Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton up to 2020

### 2016-03-02 Irregular-occ/Irregular-occ.tex
.
Feito ocultacoes 2018-2020

\title[Predictions of stellar occultations by irregular satellites up to 2020]{Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
R. Vieira-Martins$^{1,2,4,\dag}$,\newauthor
J. I. B. Camargo$^{2,4}$,
J. Desmars$^{5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{5}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{396 } % total number of occultations

@@

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}

@@

\begin{abstract}
Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument. These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations. We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future or the satellites of Jupiter. For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2017. We made observational tests for the prediction of the event of Himalia in March 03, 2015. The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions. We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017. The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
The comparisons between the predictions and the observation tests show a good agreement. We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
\end{abstract}

@@

Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}. First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software (PRAIA, \cite{Assafin2011}). Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.

@@

\begin{table*}
\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
 }\label{Tab: sat_ell}
\begin{center}
\begin{tabular}{ccccccc}
\hline\hline
Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
\hline
Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
\hline
\end{tabular} 
\end{center}
\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
\end{table*}

@@

In Fig. \ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A. 2015-Feb-27. "Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} and the \cite{Emelyanov2007} ephemeris. We can see that our ephemeris is closer to the SAT375 JPL ephemeris where the difference between them is smaller than 20 mas (< 10 mas in Declination). This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})

@@

\begin{figure*}
\begin{centering}
\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
\label{Fig:eph-Triton}
\end{centering}
\end{figure*}

@@

A total of \noccs events were identified between January 2016 and December 2017. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. No star brighter than MagR*=18 will be occulted by Triton in 2016. For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is the only found occultation by Triton and it will happen in October 05, 2017. This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
\hline
Ananke & 12 & 16 & & & &  \tabularnewline
Carme & 20 & 14 & & & &  \tabularnewline
Elara & 14 & 16 & & & &  \tabularnewline
Himalia & 15 & 12 & & & &  \tabularnewline
Leda & 8 & 24 & & & &  \tabularnewline
Lysithea & 16 & 11 & & & &  \tabularnewline
Pasiphae & 20 & 19 & & & &  \tabularnewline
Sinope & 15 & 21 & & & &  \tabularnewline
\hdashline
Phoebe & 32 & 98 & & & &  \tabularnewline
\hdashline
Nereid & 11$^{a}$ & 1 & & & &  \tabularnewline
Triton & -- & 1 & & & &  \tabularnewline
\hline
\end{tabular}
\par \end{centering}
Occultations predicted using the UCAC4 catalogue and STE ephemeris.
$^{a}$ {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
\end{table}

@@

The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016. The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris. Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be improved. Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish only the predictions for occultations restricted to the 2016-2017 time interval.

@@

We predict stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid. The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.

>>>>

\title[Predictions of stellar occultations by irregular satellites up to 2020]{Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,
R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5457 } % total number of occultations

@@

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2016}

@@

\begin{abstract}
Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument. These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations. We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future of the satellites of Jupiter. For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020. We made observational tests for the prediction of the event of Himalia in March 03, 2015. The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions. The comparisons between the predictions and the observation tests show a good agreement. We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017. The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite. We note that Jupiter will cross a populated region of stars in 2019-2020 and Saturn in 2018. This results in a large number of occultations predicted for that period.
We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
\end{abstract}

@@

Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}. First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software \citep[PRAIA,][]{Assafin2011}. Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.

@@

\begin{table*}
\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
 }\label{Tab: sat_ell}
\begin{center}
\begin{tabular}{ccccccc}
\hline\hline
Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
\hline
Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
\hline
\end{tabular} 
\end{center}
\begin{flushleft}
\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
\end{flushleft}
\end{table*}

@@

In Fig. \ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A. 2015-Feb-27. "Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} ephemeris. The difference between them is smaller than 30 mas (< 10 mas in Declination). This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})

@@

\begin{figure*}
\begin{centering}
\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
\label{Fig:eph-Triton}
\end{centering}
\end{figure*}

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. No star brighter than MagR*=18 will be occulted by Triton in 2016. For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is one of the only two found occultation by Triton and it will happen in October 05, 2017. This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
\hline
Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
\hdashline
Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
\hdashline
Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
$^{a}$ {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
\end{table}

@@

The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017). The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris. Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved. Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020. In the GAIA era, the occultation predictions will be updated.

@@

We predict stellar occultations for the period of 2016-2020 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid. The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
